Delivering Reliable Software in the European Scientific Arena: Evolution of Methodologies and New Trend Analysis

### 2018-02-07 paper-ieee-trans.tex
conclusions extended

It is a fact that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. Recent software engineering insights, such as
DevOps approaches, are gradually being applied as the open--source collaborative 
tools evolve, allowing tighter integrations. The Software Quality
Assurance procedures have been integrated in the development stage, in order to
detect and correct issues early in the software lifecycle. Automation and
metrics analysis are at the base of prompt issue solving. However, post--release
validation, through the preview testbeds or the Staged--rollout process, are
also needed to strengthen the software reliability for scientific usage.

>>>>

After more that a decade of experiences collected throughout the reviewed 
EU-funded software development and e-Infrastructure management projects, it 
becomes apparent that the European software produced for scientific purposes is
evolving to a more sustainable model where the quality and reliability of
software is being prioritized. The gradual adoption of software engineering insights, such as
DevOps practices or Agile methodologies, are leading the change. In this regard, the continuous evolution of the open--source
collaborative tools, extending their capabilities to bring along tight 
integrations amongst other services, matched the requirements of such practices
and procedures, facilitating in a great manner their implementation in real scenarios.

Reliability of the software has been substantially improved by applying the SQA
criteria in the development stage. Consequently, each change in the source code 
that is meant to be integrated in the production version, is validated according
to the SQA testing requirements. Thus, defects are detected and corrected early in the 
software lifecycle, reducing the cost of fixing errors at later stages. Automation is
key in this process, allowing the implementation of a CI scenario that enforces
the SQA criteria compliance. The application of more advanced DevOps practices
resulted in the implementation of CD pipelines to extend automation to the
delivery of packages. Thereby, safe and frequent software releases are doable, reducing
the impact on production systems. In this regard, a pre-production validation,
through the preview testbeds or resource centers early adoption -- handled by the
Staged--rollout process --, alleviates the likelihood of discovering uncovered
issues at production.

Software lifecycle demands as well from a continuous measurement and analysis
process that permits timely reactions to discover and fix defects throughout the
development, release and post-release or maintenance stages. Conscientious metric
tracking and response sits at the core of an appropriate software maturity, as 
ultimate reliability can only be achieved by enforcing and empowering the SQA 
criteria through each stage of the software lifecycle.
